\section{Conclusion and Future Work}
\label{sec:conclusion}

\subsection{Future Work}

The future work on our proposed type system will focus on the improvement of the
completeness of the type checking, and increasing the range of problems that can
be solved.

\subsubsection{Completeness Improvement:}
Our type system identify dependency of variables, by tracking the set of
variables the value of a variable depends on. Currently, we
assume that variables with the same set of dependency will have the same value,
while other cases are left until run-time to check using the run-time value. In
many cases, the value of the variable can be determined more precisely.
Another promising direction is to use the means of symbolic
execution~\cite{king1976}, where we can determine the range of vertex and edge
values that can cause conflict due to mutation.

\subsubsection{Further Expanding the Solution:}
The current type system solves a class of problems of concurrent graph mutations,
where concurrent mutation requests can conflict with each other causing
non-deterministic results or ill-formed graphs, as described in \secref{sec:motivation}. 

There are other classes of problems raised by concurrent execution of graph
mutation that is not covered by the solution proposed in this work. Consider an
algorithm where edges are added to vertices only if the out degree of a vertex
is smaller than n. Multiple attempt to add edges to a same vertex from
different vertex can be issued in the same superstep. Even though the mutation
request does not conflict with each other, the resulting graph can violate the
desired graph property. These classes of problems might be able to be solved by
expanding the expressiveness of the language to specify desired graph
properties.

\subsection{Conclusion}

The contribution of this paper is presenting a hybrid safe mutation check model for large scale graph computing that allows parallel graph mutation. Our static type checking system can type check safe mutation operations happening concurrently if the vertex identifier values and the set of dependency are known at that point. The dynamic semantics disallow the unsafe concurrent mutation operations if these can not be type checked statically. We have also implemented a prototype type checker to confirm the practicality of our approach. We envision that our approach extended with a type system that covers a wider class of problems can verify the well-formedness of graph related problems that provides the opportunity of safe graph mutation concurrently.  
