\section{A Type System with Safe Mutation}
\label{sec:types}
% More appropriate title for this section usually is "A Type System with Y"
% where Y is the new feature introduced by your approach.

We introduce two new types: vertex type and edge type. The standard arrow, pair, int, bool types are some other types in our type system. The type judgement expression consists of type environment, dependence environment, and mutation environment as shown in Figure \ref{fig-type}. Note that $\Gamma$ is a type environment mapping program variables to their annotated types; $\Lambda$ denotes the dependence environment, which stores the dependency information of a variable, i.e. a list of variables a variable depends upon before evaluation and $\Delta$ is used to denote the mutation environment that stores a mapping to dependent variables followed the mutation kind that may cause conflict associated with $\Gamma$. 

\begin{figure}[h]
\begin{center}
$\begin{array}{@{}l@{~}l@{\quad}l@{\!\!\!\!\!\!}r}
T &  ::= & \hspace{20mm} \emph{Type} \\
&\ \ \ \   \ V(T) & \hspace{20mm} \emph{Vertex Type} \\
&\ \ \ \   \mid~E(V(T),~V(T),~T) & \hspace{20mm} \emph{Edge Type} \\
&\ \ \ \   \mid~T~\Rightarrow~T & \hspace{20mm} \emph{Arrow Type} \\
&\ \ \ \   \mid~(T,~T) & \hspace{20mm} \emph{Pair Type} \\
&\ \ \ \   \mid~Int  & \hspace{20mm} \emph{Integer Type} \\
&\ \ \ \   \mid~Bool  & \hspace{20mm} \emph{Boolean Type} \\
&\ \\
&\ \\
& \textbf{Type Judgement for expressions:}
&\ \\
&\ \\
& \Gamma,~\Lambda,~\Delta~\vdash~e : T,~\Lambda,~\Delta \ \ \ \ \  where \\ 
&\ \\
& \Gamma ::= {\{ I : T_I \}}_{I \in K} & \hspace{20mm} \emph{Type Environment} \\
& \ \ \ \ \ \ \ \ where~K~is~finite \\
& \Lambda ::= {\{ x: \psi_x \}}_{x \in K} & \hspace{20mm} \emph{Dependence Environment} \\
& \ \ \ \ \ \ \ \ where~K~is~finite \\
& \psi ::= \psi ;x~\mid~\bullet & \hspace{20mm} \emph{Mutation Dependence Type} \\
& \Delta ::= \emptyset~\mid~\Delta; (\psi,~KIND); & \hspace{20mm} \emph{Mutation Environment} \\
& KIND \in \{vInsert,~vRemove,~eInsert\} & \hspace{20mm} \emph{Mutation Kinds} \\
\end{array}
$

\end{center}
\caption{Types and Type Environments.}
\label{fig-type}
\end{figure}

\subsection{Safe Mutation Check}
\label{subsec:safeMute}

The general intuition to confirm a mutation is safe i.e. conflict free is checking values of the variables in consideration and the dependency it may have on other variables. The vertices in the graph has unique identifiers. Variables have same value for every vertex unless they depend on some values that differ from vertices e.g. vertex identifier, vertex value, edge list and associated edge value. Depending on these kind of information, variables are marked with dependencies. Our current type system only keeps track of the dependency of variables, and assume that the value are the same if the dependency set is the same. Potential conflict combinations are identified as multiple vertex insertion, a vertex removal and an edge insertion issued concurrently.
\ \\
\ \\
\inferrule[\small {(Safe)}]
    {
    (KIND=vInsert~\wedge~(\nexists (\psi', vInsert) \in \Delta, \psi=\psi')~\wedge~\psi\neq \bullet )~~\vee\ \\\\
    (KIND=eInsert~\wedge~\nexists (\psi', vRemove) \in \Delta, \psi=\psi')~~\vee\ \\\\
    (KIND=vRemove~\wedge~\nexists (\psi', eInsert) \in \Delta, \psi=\psi')
    }
    {(\psi , KIND)~\#~\Delta }

\smallskip

Statically our type system checks whether two mutation expressions $e_1$ and $e_2$ happening concurrently is mutation safe ($e_1$~\#~$e_2$) or unknown at that point. The cases where it can not be known statically a runtime check is performed.

The rule (SAFE) states that to be mutation conflict safe if two vertex insert with same identifier value or vertex remove with some identifier value and an edge insertion to the vertex not with the same identifier value and vice versa are issued parallely. 

\subsection{Typing Rules for Standard Expressions}
\label{subsec:standardRules}

The typing rules presented in this section are the standard ones. Note that the mutation environment $\Delta$ do not change in any of the standard expressions. This is because we only update mutation environment when a mutation request occurs. However we update the dependency environment when a variable is concerned. 
\ \\
\ \\
\inferrule[\small {(Variable)}]
    {
    \Gamma \vdash x:T\\
    \Lambda' \vdash x:x
    }
    {\Gamma, \Lambda, \Delta \vdash x:T, \Lambda', \Delta}   

\inferrule[\small {(Value)}]
    {\ \ }
    {\Gamma, \Lambda, \Delta \vdash v:T, \Lambda, \Delta}

\inferrule[\small {(Assignment\_Int)}]
    {
     \Gamma \vdash e:Int\\
     \Gamma \vdash x:Int\\\\
     \Lambda \vdash e:\psi\\
     \Lambda' \vdash  x:\psi
     }
    {
    	\Gamma, \Lambda, \Delta \vdash x:=e : T, \Lambda ', \Delta
    }
   
\inferrule[\small {(Conditional)}]
    {
     \Gamma \vdash e_1 : Bool\\
     \Gamma, \Lambda, \Delta \vdash e_2 :T, \Lambda ', \Delta\\
     \Gamma, \Lambda, \Delta \vdash e_3 :T, \Lambda '', \Delta\\
  	 \Lambda ''' = \Lambda ' \uplus \Lambda ''
    }
    {\Gamma, \Lambda, \Delta \vdash if\ e_1 \ then\ e_2 \ else\ e_3 :T, \Lambda ''', \Delta}


\inferrule[\small {(While)}]
    {
     \Gamma \vdash e_1:Bool\\
     \Gamma, \Lambda, \Delta\vdash e_2:T, \Lambda', \Delta
    }
    {\Gamma, \Lambda, \Delta \vdash while\ e_1\ e_2:T,\Lambda', \Delta}


\inferrule[\small {(Lambda)}]
    {
     \Gamma \vdash x: T_1\\
     \Gamma[x/T_1] \vdash e: T_2}
    {\Gamma, \Lambda, \Delta \vdash \lambda x.e: T_1 \rightarrow T_2, \Lambda, \Delta}
    
 \inferrule[\small {(Application)}]
    {
     \Gamma \vdash e_1 : T_1 \rightarrow T_2\\
     \Gamma, \Lambda, \Delta \vdash e_2 : T_1, \Lambda', \Delta\\
     \Gamma, \Lambda', \Delta \vdash e_1 : T_1, \Lambda'', \Delta
     }
    {\Gamma, \Lambda, \Delta \vdash e_1 \ e_2 : T_2 , \Lambda'', \Delta}
    
 \inferrule[\small {(Operation)}]
    {
     \Gamma \vdash e_1 : T\\
     \Gamma \vdash e_2 : T\\
     \Lambda \vdash e_1 : \psi_1\\
     \Lambda \vdash e_1 : \psi_2\\
     \Lambda' = \Lambda \uplus \psi_1 \uplus \psi_2
     }
    {\Gamma, \Lambda, \Delta \vdash e_1~op~e_2 : T, \Lambda', \Delta}

\smallskip

In the (ASSIGNMENT\_INT) type rule note that dependency of the expression $e$ is same as the variable $x$. In case of (CONDITIONAL), we simply update the dependency to $\Lambda'''$ from dependency of both branches i.e. $\Lambda'$ and $\Lambda''$. The type rule for (OPERATION) is similar to conditional where dependency environment is updated based on both expressions the op is performed.\\    

 \subsection{Typing Rules for Mutation Expressions}
\label{subsec:standardRules}

Typing rules for mutation expressions are more interesting compared to other expressions. There are four possible mutation expressions in a graph based framework. They are: vertex insertion, vertex removal, edge insertion and edge removal. 
\ \\
\ \\
\inferrule[\small {(Vertex Insert)}]
    {
    \Gamma \vdash e_1 : Int\\
    \Gamma \vdash e_2 : T\\
    \Lambda \vdash e_1 : \psi\\
    \psi \# \Delta\\
    \Delta' = (\psi, vInsert) \uplus \Delta
    }
    {\Gamma, \Lambda, \Delta \vdash vInsert \ e_1 \ e_2 :Bool, \Lambda, \Delta'}

\smallskip

The rule (VERTEX INSERT) states that if a vertex insert mutation request is issued given a type environment where the vertex identifier $e_1$ is of type \emph{Int}, vertex value $e_2$ is of some type \emph{T}, the dependence environment maps the dependency of the vertex $e_1$ to be of type $\psi$ then it can be statically type checked as safe mutation if the new mutation is safe with respect to mutation environment $\Delta$ i.e. value of the vertex identifier does not depend on values stored in $\Lambda$. Then the type of this mutation is true and it updates the new mutation environment $\Delta'$ and false otherwise. It is possible to statically type check this rule only if stored mapping in $\Lambda$ are values.\\    
\ \\
\ \\    
\inferrule[\small {(Vertex Remove)}]
    {
    \Gamma \vdash e_1 : Int\\
    \Lambda \vdash e_1 : \psi\\
    \psi \# \Delta\\
    \Delta' = (\psi, vRemove) \uplus \Delta
    }
    {\Gamma, \Lambda, \Delta \vdash vRemove \ e_1 :Bool, \Lambda, \Delta'}

\smallskip

In (VERTEX REMOVE) typing rule given a vertex removal mutation and a type environment where the vertex identifier $e_1$ is of type \emph{Int} and the dependence environment maps the dependency of the vertex $e_1$ to be of type $\psi$ then it can be a safe mutation if the new mutation is safe with respect to mutation environment $\Delta$ i.e. value of the vertex identifier does not depend on values stored in $\Lambda$. Then the type of this mutation is true and it updates the new mutation environment $\Delta'$ and false otherwise. Note that it may be possible $\Lambda$ contains dependency information such that another mutation requesting adding an edge to some vertex identifier stored as value is already considered as a safe mutation and added to the mutation environment $\Delta$. By this rule if the vertex removal request has the same identifier as the identifier value associated with former mutation stored in environment, then it is possible to disallow such mutations by the static type checking system.\\    
\ \\
\ \\         
\inferrule[\small {(Edge Insert)}]
    {
    \Gamma \vdash e_1 : Int\\
    \Gamma \vdash e_2 : Int\\
    \Gamma \vdash e_3 : T\\ 
    \Lambda \vdash e_1 : \psi_1\\
    \Lambda \vdash e_2 : \psi_2\\
    \psi_1 \# \Delta\\
    \psi_2 \# \Delta\\
    \Delta' = (\psi_1, eInsert) \uplus (\psi_2, eInsert) \uplus \Delta
    }
    {\Gamma, \Lambda \Delta \vdash eInsert\ e_1\ e_2\ e_3 : Bool, \Lambda, \Delta'}

\smallskip

In the typing rule (EDGE INSERT) with an edge insert mutation and a type environment where the source vertex identifier $e_1$ is of type \emph{Int}, destination vertex identifier $e_2$ is of type \emph{Int}, edge value $e_3$ is of some type \emph{T} and the dependence environment maps the dependency of the vertex $e_1$ to be of type $\psi_1$ and vertex $e_2$ to be of $\psi_2$ then it can be a safe mutation if the new mutation is safe with respect to mutation environment $\Delta$ i.e. value of the both vertex identifiers do not depend on values stored in $\Lambda$. If this condition holds then the type of this mutation is true and it updates the mutation environment to $\Delta'$ and false otherwise.   
\ \\
\ \\ 
\inferrule[\small {(Edge Remove)}]
    {
    \Gamma \vdash e_1 : Int\\
    \Gamma \vdash e_2 : Int\\
    }
    {\Gamma, \Lambda, \Delta \vdash eRemove\ e_1\ e_2 : Bool, \Lambda, \Delta}

\smallskip

The typing rule for (EDGE REMOVE) is more straight forward as removing edges concurrently does not create any conflict or result in erroneous graph. Hence the dependence environment $\Lambda$ and mutation environment $\Delta$ remains unchanged.\\
\ \\
\ \\         
\inferrule[\small {(Sync)}]
    {\ \ }
    {\Gamma, \Lambda, \Delta \vdash sync : T, \Lambda, \Delta}
    
    \inferrule[\small {(Halt)}]
    {\ \ }
    {\Gamma, \Lambda, \Delta \vdash halt : T, \Lambda, \Delta}
    
    \inferrule[\small {(Send Message)}]
    {\Gamma \vdash e_1 :Int}
    {\Gamma, \Lambda, \Delta \vdash send\ e_1\ e_2 : Bool, \Lambda, \Delta}

 \smallskip

The typing rules for synchronization, halting computation and sending messages between vertices are simple. The dependence environment and mutation environment do not get updated. Depending on the state of these operations typing environment $\Gamma$ is updated to indicate if a specific vertex is in sync state, halt state, and message is successfully sent to destination or not.\\
\ \\
\ \\    


